COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of introduction

\section{Introduction}
In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the equation domain, since their
geometric flexibility allows solvers to be extremely effective. The
solution is sought by applying suitable numerical operations, or
kernels, to the components of a mesh, such as edges, vertices, or
cells. On standard clusters of multicores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells, possibly of the order of trillions
\cite{Rossinelli2013}, in order to obtain a satisfactory approximation
of the PDE. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centres have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office - so producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Hence, the computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper. We address, in
particular, the well-known problem of optimizing the local assembly
phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can cover a significant fraction of the overall computation run-time,
often in the range 30-60$\%$. With respect to these studies, we
propose a novel set of composable code transformations targeting, for
the first time, instruction-level parallelism - with emphasis on SIMD
vectorization - and register locality.

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in
the discretized domain. In this work, we focus on relatively low order
finite element methods, in which an assembly kernel's working set is
usually small enough to fit the L1 cache. Low order methods are by no
means exotic: they are employed in a wide variety of fields, including
climate and ocean modeling, computational fluid dynamics, and
structural mechanics. High order methods are often based upon special
formulations of the finite element method, which result in the
adoption of different sets of algorithms to solve PDEs~\cite{SPENCER},
and therefore are excluded from our study.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific,
flops-intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such a kernel structure, we focus on aspects
like minimization of floating-point operations, register allocation
and instruction-level parallelism, especially in the form of SIMD
vectorization.

Achieving high-performance is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
a single or specific sequence of transformations that is successfully
applicable to all problems. The variation in loop trip counts and
their typically small value further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized
loop-invariant code motion, vector-register tiling, and expression
splitting, as well as their composition, are essential, but none are
supported by state-of-the-art polyhedral and vendor compilers. BLAS
routines could theoretically be employed, although a fairly
complicated control- and data-flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

In order to overcome the constraints of the available compilers and
specialized linear-algebra libraries, we have automated a set of
generic and model-driven code transformations in
COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local
  assembly.}, a compiler for optimizing local assembly kernels. COFFEE
is integrated with Firedrake~\cite{firedrake-code}, a system for
solving PDEs through the finite element method based on the PyOP2
abstraction~\cite{pyop2isc}. It supports all problems expressible with
this framework, including equations that can be found at the core of
real-world simulations, such as those used in our performance
evaluation. In evaluating our code transformations in a range of
notable problems, we also vary two key parameters that impact solution
accuracy and kernel cost: the polynomial order of the method (we
investigate from $p=1$ to $p=4$) and the geometry of elements in the
discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
point of view of the generated code, these two parameters directly
impact the size of both loop nests and mathematical expressions, as
elaborated in Section~\ref{sec:real-examples}.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
remains suboptimal in the context of modern multicore
architectures. Our domian-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$. The contributions of this
paper, therefore, are
\begin{enumerate}
\item An optimization strategy for finite element local assembly. Our
  approach exploits domain knowledge and goes beyond the limits of
  both vendor and research compilers.
\item Design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item Systematic analysis using a suite of examples of real-world
  importance, and evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}

The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, showing code generated by
Firedrake and emphasizing the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain-knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. Related work is illustrated in
Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions}
reviews our contributions in the light of our results, and identifies
priorities for future work.

>>>>

\section{Introduction}

In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the computational domain, since they
allow an accurate representation of complex geometries. The
solution is sought by applying suitable numerical operations, or
kernels, to the entities of a mesh, such as edges, vertices, or
cells. On standard clusters of multicores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved by
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

% FIXME: citation for Met Office?

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions
\cite{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper. We address, in
particular, the well-known problem of optimizing the local assembly
phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can be responsible for a significant fraction of the overall computation
run-time, often in the range 30-60$\%$. With respect to these studies,
we propose a novel set of composable code transformations targeting, for
the first time, instruction-level parallelism - with emphasis on SIMD
vectorization - and register locality.

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order
finite element methods, in which an assembly kernel's working set is
usually small enough to fit the L1 cache. Low order methods are by no
means exotic: they are employed in a wide variety of fields, including
climate and ocean modeling, computational fluid dynamics, and structural
mechanics. High order methods such as the spectral element method
\cite{SPENCER} commonly require different algorithms to solve PDEs and
therefore are excluded from our study.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific,
compute intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such a kernel structure, we focus on aspects
like minimization of floating-point operations, register allocation
and instruction-level parallelism, especially in the form of SIMD
vectorization.

Achieving high-performance is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
a single or specific sequence of transformations that is successfully
applicable to all problems. Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized
loop-invariant code motion, vector-register tiling, and expression
splitting, as well as their composition, are essential. but none are
supported by state-of-the-art polyhedral and vendor compilers. BLAS
routines could theoretically be employed, although a fairly
complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

In order to overcome the constraints of the available compilers and
specialized linear algebra libraries, we have automated a set of generic
and model-driven code transformations in COFFEE\footnote{COFFEE stands
for COmpiler For FinitE Element local assembly.}, a compiler for
optimizing local assembly kernels. COFFEE is integrated with
Firedrake~\cite{firedrake-code}, a system for solving PDEs through the
finite element method based on the PyOP2
abstraction~\cite{pyop2ws,pyop2isc}. All problems expressible with this
framework are supported by COFFEE, including equations that can be found
at the core of real-world simulations, such as those used in our
performance evaluation. In evaluating our code transformations for a
range of relevant problems, we vary two key parameters that impact
solution accuracy and kernel cost: the polynomial order of the method
(we investigate from $p=1$ to $p=4$) and the geometry of elements in the
discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
point of view of the generated code, these two parameters directly
impact the size of both loop nests and mathematical expressions, as
elaborated in Section~\ref{sec:real-examples}.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
% FIXME: what are these state-of-the-art techniques?
remains suboptimal in the context of modern multicore
architectures. Our domain-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$. The contributions of this
paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}

The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. Related work is described in
Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions}
reviews our contributions in the light of our results, and identifies
priorities for future work.
